Capturing the geometric structure of our experiences and how we remember them

### 2018-08-16 paper/main.tex
started discussion

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our experiences is that they unfold in a structured and predictable manner, governed by the laws of the physical systems that produce them. However, this temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the inherent temporal structure of our life experiences influences memorability and memory organization is limited. Using a novel computational modelling approach, we investigate how people verbally recall continuous videos by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. We found that despite large differences in the quantity of information recalled across participants, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

@@

>>>>

\section{Abstract}
{
The human memory system is adept at cataloging the rich dynamics of ongoing experience. A defining feature of our everyday experiences is that they unfold in a structured and predictable manner, governed by the laws of the physical systems that produce them. However, this temporal structure is typically absent (or severly impoverished) in traditional trial-based memory experiments. As a consequence, our understanding of how the inherent temporal structure of our experiences influences memorability and memory organization is limited. In this study, we investigate how people verbally recount a video by characterizing and relating the thematic dynamics, or ``trajectories,'' of the stimulus and participants' recalls. We found that despite large differences in the amount of information recalled across participants as well as the particular words used to describe the experience, the overall shape of the stimulus was recapitulated by most of the participants. These findings provide a window into which aspects of naturalistic experiences must be preserved, and which might be more flexible, in considering whether and how those experiences are remembered.

@@

Studying episodic memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall and item recognition experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. However, this approach does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, allows us to flexibly and accurately characterize memory for naturalistic experiences. This allows us to quantify which moments from the past the current thoughts match, in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words.

Historically, human memory reseachers have focussed on measuring the quantity of information recovered by an individual under different experimental conditions (i.e. the proportion correct). However, for real world experiences this is not a well-defined task. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience. Our work characterizes experience by comparing the overall ``shape'' of an experience and a memory. We assess the quality of memory for the movie participants viewed by measuring the match between the shapes of the movie's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory-- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape.

Prior work on neural responses during naturalistic experience and recall (e.g. Hasson group) has largely focused on identifying brain regions whose responses are reliably similar across individuals. This allows one to identify which regions might be processing or representing the stimulus, or retrieving details of the experience during recall. Our approach is fundamentally different. We ask: which brain regions during movie viewing match how individual participants will recall the movie later (in terms of the temporal correlations of their recall topic trajectory)? This approach highlights regions such as the vmPFC that may respond in idiosyncratic ways across individuals, but that nonetheless play an important role in encoding experience into memory.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, like the in the classroom or in a doctor's office.  Whereas instructors often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might miss the ``forrest for the trees''. True learning is about understanding the key concepts (i.e. understanding key themes in the learned content and how they relate) rather than about regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of a memory disorder, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the movie's topic trajectory? Or if participants were told that their narrations would be played back to other participants (cite Hasson work), would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing our" of the trajectories with each new retelling. This could be quantified using our topic trajectory approach.

In summary, using a novel content-based memory assessment approach,
